---
title: 'Making Health Economics Shiny: A tutorial'
author: "Robert Smith, Paul Schneider & Sarah Bates"
date: "18/02/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

R Markdown documents can be used to write reports. In this example I am rendering this report to PDF, but it could be output in Word or as an online document (like the course material).

# The Sick Sicker Model Results

```{r, echo = F}

library(ggplot2)
library(truncnorm)

# define folders (relative not working)
course_folder = "C:/Users/Robert/Google Drive/Dark Peak Analytics/emsr_course"
src_folder = "Markov Model/src"
plot_folder = "Advanced R Shiny/HE_Plots"

# source the scripts that contain the model functions from the Markov Model/src folder.
source(paste0(course_folder, "/", src_folder, "/", "f_gen_psa.R"))
source(paste0(course_folder, "/", src_folder,"/", "f_MM_sicksicker.R"))
source(paste0(course_folder, "/", src_folder,"/", "f_wrapper.R"))
source(paste0(course_folder, "/", plot_folder,"/","plotCEAC.R"))
source(paste0(course_folder, "/", plot_folder,"/","plotCEP.R"))

params <- list(
  initial_age = 20,
  maximum_age = 75,
  discount_rate = 0.035,
  n_iterations = 1000,
  drug_price = 200)

```
We can run the entire model from within Rmarkdown, and produce outputs:

```{r, echo=TRUE, message=FALSE}

df_results <- f_wrapper(n_age_init = params$drug_price,
                        n_age_max  = params$n_iterations,
                        d_r        = params$discount_rate,
                        n_sim      = params$n_iterations,
                        c_Trt      = params$drug_price)

head(df_results)

```


## Including Plots

You can also embed plots, for example here is a the Cost Effectiveness Plane.

```{r, echo=TRUE, message=FALSE}

plotCEP(df_results)

```

```{r, echo=FALSE, message=FALSE}

ICER <- sum(df_results$Cost_Trt - df_results$Cost_NoTrt) / sum(df_results$QALY_Trt - df_results$QALY_NoTrt)

```

Note that the text can include model results. For example I could say that the ICER is 
`r paste0("£", round(ICER,2))`

# Exercise

1. Your client has mentioned that their treatment cost is now £375, rather than the £200 stated in the previous report. They would like to see an identical report to that shown above, but with an updated cost-effectiveness plane and ICER.

2. Your client would like to see the CEAC curve and for the insertion of a paragraph at the beginning with the inputs to the Sick-Sicker function defined. (Ensure you do this so that the entire document can be updated by one change).

3. The decision body has changed its discount rate from 3.5% to 1.5%. Please update the entire report.

