---
title: "Partitioned Survival Models (PSM) R tutorial "
output: html_notebook
---


# Recap

  * In a Partitioned Survival Model (PSM), individuals reside in one of a series of mutually exclusive and jointly exhaustive health states. 
  
  * State membership is usually determined by a series of independently modelled non-mutually exclusive survival curves. 
  
  * A survival curve must be specified for each alive health state that describes time from model start until transiting to another health state.
  
  * Evaluating PSM is different to Markov models. Health state occupancy ( *sojourn*) time is estimated directly from curves (as integral/AUC), rather than as function of transition matrices.
  
  

# TUTORIAL

  Suppose you have been asked to evaluate a new treatment 'Supimab', against the standard-of-care treatment (SOC). The model has the following specifications:

Treatment options

  * <b>SUPIMAP</b>
  * <b>STANDARD-OF-CARE (SOC)</b>


Model type: PSM with three states

  * PFS = Progression-free survival
  * PPS = Post-progression survival
  * dead
  (N.B.: OS = overall survival = PFS + PPS = 1 - dead)

Patients in PFS receive SUPIMAB or SOC, after transitioning to PPS, all patients 'supportive care'.


```{r echo=FALSE}
library(knitr)
library(kableExtra)

tbl <- t(cbind(

# suoimab
"costs drug" = 45000,
"costs administration" = 2000,
"OS (Weibull):"="",
"constant" = 0.3074,
"scale" = 0.3536,
"PFS (Weibull):"="",
"constant" = 0.9290,
"scale" = 0.4408,


# soc
"drug costs" = 22000,
"OS (Weibull)" ="",
"constant" = 0.4583,
"scale" = 0.3249,
"PFS (Weibull)" = "",
"constant" = 1.8498,
"scale" = 0.4337,

# general
"HRQoL PFS" = 0.90,
"HRQoL PPS"  = 0.50,
"costs PPS" = 5000,
'Time horizon' = "10 years",
'Discount rate' = "3.5%"
))
kable(tbl,caption = "Parameters") %>%
  kable_paper(full_width = F) %>%
    group_rows(group_label = "Supimab",start_row = 1,end_row = 8) %>% 
  group_rows(group_label = "SOC",start_row = 9,end_row = 15) %>% 
  group_rows(group_label = "Other",start_row = 16,end_row = 20)

```


## Setup

```{r}
# load packages
  # install.packages("flexsurv")
  # install.packages("ggplot2")
  library(flexsurv)
  library(ggplot2)

# set horizon and time intervals
  time_steps <- seq(from = 0, to = 10, by = 0.5)
  # use 0.5 to illustrate model
  # use <0.01 for higher accuracy


# S(t) Weibull function
  stWB <- function(const,scale,time){
    res <- exp( -const * (time^exp(scale)) )
    return(res)
  }

```



```{r}
# SUPIMAB survival curves -------

  # OS
  os_supi_wb_const <- 0.3074
  os_supi_wb_scale <- 0.3536
  St_supi_os <-  stWB(os_supi_wb_const,  os_supi_wb_scale,  time_steps)

  # PFS
  pfs_supi_wb_const <- 0.9290
  pfs_supi_wb_scale <- 0.4408
  St_supi_pfs <- stWB(pfs_supi_wb_const, pfs_supi_wb_scale, time_steps)
  
  # PPS = OS - PFS
  St_supi_pps <- St_supi_os - St_supi_pfs
  
  # dead = 1 - OS
  St_supi_dead <- 1 - St_supi_os
  
  
# SOC survival curves -------
  
  # OS
  os_soc_wb_const <- 0.4583
  os_soc_wb_scale <- 0.3249
  St_soc_os <-  stWB(os_soc_wb_const,  os_soc_wb_scale,  time_steps)
  
  # PFS
  pfs_soc_wb_const <- 1.8498
  pfs_soc_wb_scale <- 0.4337
  St_soc_pfs <- stWB(pfs_soc_wb_const, pfs_soc_wb_scale, time_steps)
  
  # PPS = OS - PFS
  St_soc_pps <- St_soc_os - St_soc_pfs

  # dead = 1 - OS
  St_soc_dead <- 1 - St_soc_os
```


# Visualise fitted survival curves with ggplot2

```{r}
  # proportion in state
    pm_supi <- ggplot() +
      geom_line(aes(x= time_steps, y = St_supi_os, col = "OS")) +
      geom_line(aes(x= time_steps, y = St_supi_pfs, col = "PFS")) +
      geom_line(aes(x= time_steps, y = St_supi_pps, col = "PPS")) +
      geom_line(aes(x= time_steps, y = St_supi_dead, col = "dead"))   +
      ggtitle("Supimab") +
      ylab("Proportion in state") +
      theme_minimal()
    
  pm_supi

  
    # time spend in each state
    pm_soc <-  ggplot() +
      geom_line(aes(x= time_steps, y = St_soc_os, col = "OS")) +
      geom_line(aes(x= time_steps, y = St_soc_pfs, col = "PFS")) +
      geom_line(aes(x= time_steps, y = St_soc_pps, col = "PPS")) +
      geom_line(aes(x= time_steps, y = St_soc_dead, col = "dead"))   +
      ggtitle("SOC") +
      ylab("Proportion in state") +
      theme_minimal()

    pm_soc
    
    cowplot::plot_grid(pm_supi ,pm_soc, nrow = 2)  
  


# visualise fitted survival curves
  ggplot() +
    # SOC
    geom_point(aes(x = time_steps, y = St_soc_pfs, col = "SOC")) +
    geom_line(aes(x = time_steps, y = St_soc_pfs, linetype = "PFS", col = "SOC")) +
    geom_point(aes(x = time_steps, y = St_soc_os, col = "SOC")) +
    geom_line(aes(x = time_steps, y = St_soc_os, linetype = "OS",col = "SOC")) +
    # supimab
    geom_point(aes(x = time_steps, y = St_supi_pfs, col = "Supimab")) +
    geom_line(aes(x = time_steps, y = St_supi_pfs, linetype = "PFS",col = "Supimab")) +
    geom_point(aes(x = time_steps, y = St_supi_os, col = "Supimab")) +
    geom_line(aes(x = time_steps, y = St_supi_os, linetype = "OS",col = "Supimab")) +
    # labels, axes, and legend
    ggtitle("Supimab and SOC survival curves") +
    ylab("Cumulative survival") +
    scale_x_continuous(name = "Years", breaks = seq(0,10,2)) +
    scale_color_manual(name = "Treamtnet", 
                       labels = c("SOC", "Supimab"),
                       values = c("#F8766D","#00BFC4")) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  # visualise AUC PFS Supimab
  ggplot() +
    geom_ribbon(aes(x = time_steps, ymin = 0,ymax = St_supi_pfs, fill = "PFS")) +
    geom_line(aes(x = time_steps, y = St_supi_pfs)) +
    geom_line(aes(x = time_steps, y = St_supi_os)) +
    # labels, axes, and legend
    ggtitle("PFS AUC - Supimab ") +
    ylab("Cumulative survival") +
    scale_x_continuous(name = "Years", breaks = seq(0,10,2)) +
    scale_fill_manual(name = "State", 
                       labels = c("PFS AUC"),
                       values = c("#00BFC4")) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  # visualise AUC PPS + PFS Supimab
  ggplot() +
    geom_ribbon(aes(x = time_steps, ymin = 0,ymax = St_supi_pfs, fill = "PFS")) +
    geom_ribbon(aes(x = time_steps, ymin = St_supi_pfs,ymax = St_supi_os, fill = "PPS"), alpha = 0.7) +
    geom_line(aes(x = time_steps, y = St_supi_pfs)) +
    geom_line(aes(x = time_steps, y = St_supi_os)) +
    # labels, axes, and legend
    ggtitle("PFS+PPS AUC - Supimab ") +
    ylab("Cumulative survival") +
    scale_x_continuous(name = "Years", breaks = seq(0,10,2)) +
    scale_fill_manual(name = "State", 
                       labels = c("PFS AUC","PPS AUC"),
                       values = c("#00BFC4","cadetblue")) +
    theme_minimal() +
    theme(legend.position = "bottom")
```


## Compute AUC

> trapezium rule
> AUC = ( (y2+y1)/2)*(x2 - x1)

```{r echo=FALSE}
# trapeziums illustration
vline_df <- data.frame(cbind(
  x = rep(time_steps,2),
  y = c(cbind(0,St_supi_pfs)),
  id = as.factor(rep(time_steps,2))
  ))

ggplot() +
    geom_ribbon(aes(x = time_steps[2:3], ymin = 0,ymax = St_supi_pfs[2:3], fill = "PFS")) +
    geom_line(aes(x = time_steps, y = St_supi_pfs)) +
    geom_line(aes(x = time_steps, y = St_supi_pfs)) +
    geom_line(data = vline_df, aes(x = x, y = y,group = id), linetype = "dashed") +
    geom_point(aes(
        x=c(time_steps[2:3],time_steps[2:3]),
        y=c(0,0,St_supi_pfs[2:3]))) +
    ggrepel::geom_label_repel(aes(
        x=c(time_steps[2:3],time_steps[2:3]),
        y=c(0,0,St_supi_pfs[2:3]),
        label = c("x1","x2","y1","y2")
        )) +
    geom_line(aes(
      x=c(0,time_steps[2]),
      y=c(St_supi_pfs[2],St_supi_pfs[2])
      ), linetype = "dotted") +
    geom_line(aes(
      x=c(0,time_steps[3]),
      y=c(St_supi_pfs[3],St_supi_pfs[3])
      ), linetype = "dotted") +
    theme_minimal() +
    xlim(c(0,4)) +
    geom_text(aes(x = 2.5,y = 0.65,label = "AUC = ( (y2+y1)/2)*(x2 - x1)"),size = 10) +
    theme(legend.position = "none")
```


```{r}
# example for one trapezium
((St_supi_pps[3] + St_supi_pps[2]) / 2 ) * (time_steps[3] - time_steps[2])
```


```{r}
# build your own AUC function
# AUC = ((y2+y1)/2)*(x2 - x1)
myAUC <- function(x1,x2,y1,y2){
  res <- ((y2+y1)/2)*(x2 - x1)
  return(res)
}

# loop 1 - through 1 St obj
mat_auc <- c()
for(t in 1:(length(time_steps) - 1 )){   # lenght - 1 to avoid NA 
  auc_pps <- myAUC(
    x1 = time_steps[t],
    x2 = time_steps[t+1],
    y1 = St_supi_pps[t],
    y2 = St_supi_pps[t+1]
    )
  mat_auc <- rbind(mat_auc,auc_pps)
}

head(mat_auc)
# tail(mat_auc)
sum(mat_auc)

# side step:
  # we could have used a package instead
  # install.packages("DescTools")
  library(DescTools)
  ?AUC
  AUC # much shorter, but hard to read code
  AUC(x=time_steps, y=St_supi_pps, method="trapezoid")

  
  
# AUC for all states:
# combine vectors of cumulative survival probabilities
mat_surv <- cbind(
  St_supi_pfs,St_supi_pps,St_supi_dead,
  St_soc_pfs,St_soc_pps,St_soc_dead
  )

mat_surv

mat_auc <- c()
for(i in 1:ncol(mat_surv)){
  temp_auc <-c()
  for(t in 1:(length(time_steps) - 1 )){   # lenght - 1 to avoid NA 
    temp_auc_step <- myAUC(
      x1 = time_steps[t],
      x2 = time_steps[t+1], 
      y1 = mat_surv[,i][t],
      y2 = mat_surv[,i][t+1]
      )
    temp_auc <- c(temp_auc, temp_auc_step)
  }
  mat_auc <- cbind(mat_auc,temp_auc)
}

colnames(mat_auc) <- colnames(mat_surv)
head(mat_auc)

auc_df <- reshape2::melt(mat_auc)
names(auc_df) <- c("time","state","auc")
auc_df$trt <- grepl("supi", auc_df$state)
auc_df$state <- gsub("soc_","",auc_df$state)
auc_df$state <- gsub("supi_","",auc_df$state)

ggplot(auc_df, aes(fill=state, y=auc, x=time)) + 
    geom_bar(position="stack", stat="identity") +
    facet_wrap(~trt) +
    theme_minimal()

sum(mat_auc) # total auc check
rowSums(mat_auc) # time stpes check
colSums(mat_auc) # col auc sums
```




## HRQoL and costs
```{r}
# utilities
u_pfs <- 0.90
u_pps <-  0.50

# costs soc
c_PFS_soc <- 22000

# costs supi
c_drug_supi <- 45000
c_admin_supi <- 2000
c_PFS_supi <- c_drug_supi + c_admin_supi

# costs supportive care for post-progression
c_support <- 5000
```


## outcomes
```{r}

# head(mat_auc)

c_pfs_supi <- mat_auc[,"St_supi_pfs"] * (c_PFS_supi)
c_pps_supi <- mat_auc[,"St_supi_pps"] * c_support
c_pfs_soc <- mat_auc[,"St_soc_pfs"] * (c_PFS_soc)
c_pps_soc <- mat_auc[,"St_soc_pps"] * c_support

q_pfs_supi <- mat_auc[,"St_supi_pfs"] * u_pfs
q_pps_supi <- mat_auc[,"St_supi_pps"] * u_pfs
q_pfs_soc <- mat_auc[,"St_soc_pfs"] * u_pfs
q_pps_soc <- mat_auc[,"St_soc_pps"] * u_pps

# # shoter
# c_vector <- c(c_PFS_supi, c_support, 0, c_PFS_soc, c_support, 0)
# t(apply(mat_auc,1,function(x)  x* c_vector))
# u_vector <- rep(c(u_pfs,u_pps,0),2)
# t(apply(mat_auc,1,function(x)  x* u_vector))

c_mat <- cbind(
  c_pfs_supi,c_pps_supi, 
  c_pfs_soc, c_pps_soc
  )

q_mat <- cbind(
  q_pfs_supi,q_pps_supi,
  q_pfs_soc,q_pps_soc
)
```

# discounting
x /(1 + disc_rate) ^ time
e.g. cost /(1.035) ^ 0.05

```{r}
myDiscounter <- function(x, time, disc_rate = 0.035){
 res <- x /(1 + disc_rate) ^ time 
 return(res)
}

# c_pfs_supi_disc <-  myDiscounter(x = c_pfs_supi,time = time_steps[-1])
# head(c_pfs_supi_disc)

c_mat_disc <- apply(c_mat,2,myDiscounter,time = time_steps[-1])
head(c_mat_disc)

q_mat_disc <- apply(q_mat,2,myDiscounter,time = time_steps[-1])
head(q_mat_disc)
```



# Results

```{r}
# costs supimab
tc_supi <- sum(c_mat_disc[,c("c_pfs_supi","c_pps_supi")])
tc_supi
# qalys supimab
tq_supi <-sum(q_mat_disc[,c("q_pfs_supi","q_pps_supi")])
tq_supi
# costs soc
tc_soc <- sum(c_mat_disc[,c("c_pfs_soc","c_pps_soc")])
tc_soc
# qalys soc
tq_soc <- sum(q_mat_disc[,c("q_pfs_soc","q_pps_soc")])
tq_soc

# ICER
mean( (tc_supi - tc_soc) / (tq_supi - tq_soc) )

# INB at 20k
inb <-  (tq_supi - tq_soc) * 20000 - (tc_supi - tc_soc)
inb

# INB at 20k
inb <-  (tq_supi - tq_soc) * 20000 - (tc_supi - tc_soc)
inb
```
<br><br>

*****

<br><br>

### Additional exercises:

1. Build a function `calcInb` to compute the INB for any wtp
  - the function should task as arguments the total costs and qalys of two different treatments
  and a threshold value, and return the INB. (set the default value to 20,000)
  
2. calculate the inb using the following parameters:

  threshold values = 0, 5,000, 10,000, 15,000, ..., 50,000

  *treatment 1*
  costs = 48243
  qalys = 1.8
  *treatment 2*
  costs = 17779
  qalys = 1
  
3. create a ggplot2 that shows inb on the y axis and the thresholds on the x-axis
  - plot a line and points

```{r}
calcInb <- function(c_1,c_2,q_1,q_2,wtp = 20000){
  res <-  (q_1 - q_2) * wtp - (c_1 - c_2)
  res
}

thres_vals <- seq(from = 0, to = 50000,by = 5000)

inb_vals <- calcInb(c_1 = 48243, q_1 = 1.8, c_2 = 17779,q_2 = 1,wtp = thres_vals)

ggplot() +
  geom_point(aes(x = thres_vals, y = inb_vals)) +
  geom_line(aes(x = thres_vals, y = inb_vals))  +
  geom_hline(yintercept = 0)
```
