---
title: 'Partitioned Survival Models (PSM) R tutorial '
output:
  html_document:
    df_print: paged
---


### Recap

  * In a Partitioned Survival Model (PSM), individuals reside in one of a series of mutually exclusive and jointly exhaustive health states. 
  
  * State membership is usually determined by a series of independently modelled non-mutually exclusive survival curves. 
  
  * A survival curve must be specified for each alive health state that describes time from model start until transiting to another health state.
  
  * Evaluating PSM is different to Markov models. Health state occupancy ( *sojourn*) time is estimated directly from curves (as integral/AUC), rather than as function of transition matrices.
  
  

### TUTORIAL

  Suppose you have been asked to evaluate a new treatment 'Supimab', against the standard-of-care treatment (SOC). The model has the following specifications:

Treatment options

  * <b>SUPIMAP</b>
  * <b>STANDARD-OF-CARE (SOC)</b>


Model type: PSM with three states

  * PFS = Progression-free survival
  * PPS = Post-progression survival (= OS - PFS)
  * dead (= 1 -  OS)
  (N.B.: OS = overall survival = PFS + PPS)

Patients in PFS receive either supimab or SOC. After disease progression, all patients stop their treatment and receive supportive care.


### Setup

```{r}
  options(scipen=99, digits=3)

## load packages
  library(knitr)
  library(survival)
  library(flexsurv)
  library(ggplot2)
```


### Preamble 

**Fitting parametric survival models and retrieving parameters for the PSM**

  Suppose we have some individual patient data from a clinical trial, in which 200 participants were randomised to either receiving standard of care or the new drug Supimab (100 to each group). In the trial, the time until progression and the time until death were recorded. For simplicity, there will be no censoring of data. (The individual patient data was simulated using the simsurv package - see [source code](https://github.com/bitowaqr/emsr_course/blob/main/PSM/sim_surv.R))
  
  We will (1) load the data sets, (2) fit a Weibull model (PH) to both OS and PFS data, and then (3) retrieve the shape and scale parameters, which we will use in the PSM model.

```{r}
## Overall survival data
# 1 load data and fit parametric models
  df_os <- read.csv("../PSM/data/trial_os.csv")
  head(df_os)
  
# 2 fit a paraetric survival model
  surv_os <- Surv(time = df_os$eventtime, event = df_os$status) 
  os_reg <- flexsurvreg(surv_os ~ trt, data = df_os,dist = "weibullph")
  os_reg

# 3 retrieve parameters from models 
  os_shape <- os_reg$res[1,1]
  os_shape
  os_scale_soc <- os_reg$res[2,1]
  os_scale_soc
  os_scale_supi <- os_reg$res[2,1] * exp(os_reg$res[3,1])
  os_scale_supi
  
## repeat the same for PFS
## Progression-free survival 
# 1 load data and fit parametric models
  df_pfs <- read.csv("../PSM/data/trial_pfs.csv")
  head(df_pfs)
  
# 2 fit a paraetric survival model
  surv_pfs <- Surv(time = df_pfs$eventtime, event = df_pfs$status)
  pfs_reg <- flexsurvreg(surv_pfs ~ trt, data = df_pfs,dist = "weibullph")
  pfs_reg

# 3 retrieve parameters from models   
  pfs_shape <- pfs_reg$res[1,1]
  pfs_shape
  pfs_scale_soc <- pfs_reg$res[2,1]
  pfs_scale_soc
  pfs_scale_supi <- pfs_reg$res[2,1] * exp(pfs_reg$res[3,1])
  pfs_scale_supi
  
```


### Define model parameters

  

```{r echo=FALSE}
library(knitr)
library(kableExtra)

tbl <- t(cbind(

## suoimab
"costs drug" = 30000,
"costs administration" = 2000,
"OS (Weibull):"="",
"shape" = 1.530795,
"scale" = 0.2734298,
"PFS (Weibull):"="",
"shape" = 1.543373,
"scale" = 0.8797236,


## soc
"drug costs" = 22000,
"OS (Weibull)" ="",
"shape" = 1.530795,
"scale" = 0.5459197,
"PFS (Weibull)" = "",
"shape" = 1.543373,
"scale" = 2.389363,

## general
"utility PFS" = 0.90,
"utility PPS"  = 0.65,
"costs PPS" = 5000,
'Time horizon' = "10 years",
'Discount rate' = "3.5%"
))
kable(tbl,caption = "Parameters") %>%
  kable_paper(full_width = F) %>%
    group_rows(group_label = "Supimab",start_row = 1,end_row = 8) %>% 
  group_rows(group_label = "SOC",start_row = 9,end_row = 15) %>% 
  group_rows(group_label = "Other",start_row = 16,end_row = 20)

```

To illustrate how the PSM works, we will use very large time intervals of one year - in practice, the interval should be much smaller, i.e. 1 day.

```{r}
  # Define horizon and time intervals 
  times <- seq(from = 0, to = 10, by = 1) # NOTE: 1 interval = 1 year 
  
  # utilities
  u_pfs <- 0.90
  u_pps <-  0.65
  
  # costs soc
  c_soc <- 22000
  
  # costs supimab
  c_drug_supi <- 30000
  c_admin_supi <- 2000
  c_supi <- c_drug_supi + c_admin_supi
  
  # costs supportive care for post-progression
  c_support <- 5000
  
  # discount rate
  disc_rate = 0.035
```

### Predict cumulative survival curves for OS, PFS, PPS & Dead

```{r}
  ## OS curves
  pred_os_soc <- 1 - os_reg$dfns$p(times,shape=os_shape,scale = os_scale_soc)
  pred_os_supi <- (1 - os_reg$dfns$p(times,shape=os_shape, scale = os_scale_supi))
  
  ## PFS curves
  pred_pfs_soc <- 1 - pfs_reg$dfns$p(times,shape=pfs_shape,scale = pfs_scale_soc)
  pred_pfs_supi <- (1 - pfs_reg$dfns$p(times,shape=pfs_shape, scale = pfs_scale_supi))
  
  ## PPS curves
  pred_pps_soc <- pred_os_soc - pred_pfs_soc
  pred_pps_supi <- pred_os_supi - pred_pfs_supi
  
  ## dead curves
  pred_dead_soc <- 1 - pred_os_soc
  pred_dead_supi <- 1 - pred_os_supi

  # Combine estimated survival curves into a matrix
  mat_surv <- cbind(
    pred_pfs_soc, pred_pps_soc, pred_dead_soc,
    pred_pfs_supi, pred_pps_supi, pred_dead_supi
  )
  kable(mat_surv,digits = 3)
```


### Visualise fitted survival curves

```{r}
  ### visual inpsection of curves ------
  kmPlotter <- function(event_times, group){
    # simplified KM plot df function
    # can only be used when there is NO CENSORING!
    uniq_g <- unique(group)
    res<- c()
    for(g in uniq_g){
      times <- event_times[group ==g]
      min_t <- min(times)
      max_t <- max(times)
      uniq_t <- unique(times)
      uniq_t <- uniq_t[order(uniq_t)]
      n <- length(times)
      S <- c(1)
      for(t in 2:length(uniq_t)){
        post_n <- sum(times > uniq_t[t])
        S <- c(S,post_n/ n  )
      }
      df <- data.frame(times = uniq_t, S, trt = g) 
      res <- rbind(res,df)
    }
    
    return(res)
  }
  
  km_plot_os <- kmPlotter(df_os$eventtime,df_os$trt)
  km_plot_pfs <- kmPlotter(df_pfs$eventtime,df_pfs$trt)
  
  
  # PLOT OBSERVED AND PREDICTED TOGETHER
  ggplot() +
    # OS KM observed
    geom_step(data = km_plot_os,aes(x=times, y= S, col = trt, linetype="OS"), alpha = 0.5) +
    # PFS KM observed
    geom_step(data = km_plot_pfs,aes(x=times, y= S, col = trt, linetype="PFS"), alpha = 0.5) +
    
    # Weibull est OS
    geom_line(aes(x=times,y=pred_os_soc,col = "SOC", linetype="OS" )) +
    geom_line(aes(x=times,y=pred_os_supi, col="Supimab",linetype="OS" )) +
    # Weibull est PFS
    geom_line(aes(x=times,y=pred_pfs_soc,col = "SOC", linetype="PFS" )) +
    geom_line(aes(x=times,y=pred_pfs_supi, col="Supimab",linetype="PFS" )) +
    ggtitle("Observed (KM) and predicted (Weibull) OS and PFS curves") +
    coord_cartesian(xlim=c(0,7)) +
    theme_minimal()
  
  # AUC
  ## visualise AUC PPS + PFS Supimab
  ggplot() +
    geom_ribbon(aes(x = times, ymin = 0,ymax = pred_pfs_supi, fill = "PFS")) +
    geom_ribbon(aes(x = times, ymin = pred_pfs_supi,ymax = pred_os_supi, fill = "PPS"), alpha = 0.7) +
    geom_line(aes(x = times, y = pred_pfs_supi)) +
    geom_line(aes(x = times, y = pred_os_supi)) +
    ## labels, axes, and legend
    ggtitle("PFS+PPS AUC - Supimab ") +
    ylab("Cumulative survival") +
    scale_x_continuous(name = "Years", breaks = seq(0,10,2)) +
    scale_fill_manual(name = "State", 
                      labels = c("PFS AUC","PPS AUC"),
                      values = c("#00BFC4","cadetblue")) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  ## visualise AUC PPS + PFS SOC
  ggplot() +
    geom_ribbon(aes(x = times, ymin = 0,ymax = pred_pfs_soc, fill = "PFS")) +
    geom_ribbon(aes(x = times, ymin = pred_pfs_soc,ymax = pred_os_soc, fill = "PPS"), alpha = 0.7) +
    geom_line(aes(x = times, y = pred_pfs_soc)) +
    geom_line(aes(x = times, y = pred_os_soc)) +
    ## labels, axes, and legend
    ggtitle("PFS+PPS AUC - SOC ") +
    ylab("Cumulative survival") +
    scale_x_continuous(name = "Years", breaks = seq(0,10,2)) +
    scale_fill_manual(name = "State", 
                      labels = c("PFS AUC","PPS AUC"),
                      values = c("#00BFC4","cadetblue")) +
    theme_minimal() +
    theme(legend.position = "bottom")
```


### Compute AUCs

> trapezium rule
> AUC = ( (y2+y1)/2)*(x2 - x1)

```{r echo=FALSE}
## trapeziums illustration
vline_df <- data.frame(cbind(
      x = rep(times,2),
      y = c(cbind(0,pred_pfs_supi)),
      id = as.factor(rep(times,2))
    ))
    
    ggplot() +
      geom_ribbon(aes(x = times[2:3], ymin = 0,ymax = pred_pfs_supi[2:3], fill = "PFS")) +
      geom_line(aes(x = times, y = pred_pfs_supi)) +
      geom_line(data = vline_df, aes(x = x, y = y,group = id), linetype = "dashed") +
      geom_point(aes(
        x=c(times[2:3],times[2:3]),
        y=c(0,0,pred_pfs_supi[2:3]))) +
      ggrepel::geom_label_repel(aes(
        x=c(times[2:3],times[2:3]),
        y=c(0,0,pred_pfs_supi[2:3]),
        label = c("x1","x2","y1","y2")
      )) +
      geom_line(aes(
        x=c(0,times[2]),
        y=c(pred_pfs_supi[2],pred_pfs_supi[2])
      ), linetype = "dotted") +
      geom_line(aes(
        x=c(0,times[3]),
        y=c(pred_pfs_supi[3],pred_pfs_supi[3])
      ), linetype = "dotted") +
      theme_minimal() +
      xlim(c(0,4)) +
      geom_text(aes(x = 2.5,y = 0.65,label = "AUC = ( (y2+y1)/2)*(x2 - x1)"),size = 10) +
      theme(legend.position = "none") +
      ylab("S(t)") +
      xlab("Years")
```


```{r}
# define and apply AUC function
  myAUC <- function(x, y){
    auc <- rep(NA, times = length(x)-1)
    for (i in 2:(length(x))) {
      auc_i <- ( (y[i] + y[i - 1])/2 ) * (x[i] - x[i - 1])
      auc[i-1] <- auc_i
    }
    return(auc)
  }
  
  mat_auc <- apply(mat_surv,2,myAUC,x = times)
  kable(mat_auc,digits = 3)

  ## Run some basic checks  
  rowSums(mat_auc) # time stepes check
  sum(mat_auc) # total auc check
  colSums(mat_auc) # time spend in state
```
  


### Multiply costs and utilities with computed AUCs
```{r}
  c_pfs_supi <- mat_auc[,"pred_pfs_supi"] * c_supi
  c_pps_supi <- mat_auc[,"pred_pps_supi"] * c_support
  c_pfs_soc <- mat_auc[,"pred_pfs_soc"] * c_soc
  c_pps_soc <- mat_auc[,"pred_pps_soc"] * c_support
  
  q_pfs_supi <- mat_auc[,"pred_pfs_supi"] * u_pfs
  q_pps_supi <- mat_auc[,"pred_pps_supi"] * u_pps
  q_pfs_soc <- mat_auc[,"pred_pfs_soc"] * u_pfs
  q_pps_soc <- mat_auc[,"pred_pps_soc"] * u_pps

  # combines costs and qalys into matrices
  # costs 
  c_mat <- cbind(c_pfs_supi, c_pps_supi, c_pfs_soc, c_pps_soc)
  kable(c_mat,digits = 2)
  
  # qalys
  q_mat <- cbind(q_pfs_supi, q_pps_supi, q_pfs_soc, q_pps_soc)
  kable(q_mat, digits = 3)
```


### Define discount function and apply it to costs and qalys

  Discounting:
  x /(1 + disc_rate) ^ time
  e.g. cost /(1.035) ^ 0.05

```{r}
  myDiscounter <- function(x, time, rate = 0.035){
    res <- x /(1 + rate) ^ time 
    return(res)
  }
  
  # apply discounting to cost matrix
  c_mat_disc <- apply(c_mat, 2, myDiscounter, time = times[-1])
  kable(c_mat_disc,digits = 2)
  
  # apply discounting to qaly matrix
  q_mat_disc <- apply(q_mat,2,myDiscounter,time = times[-1])
  kable(q_mat_disc, digits = 3)
```


### Sum total (discounted) costs and qalys and compute ICER + INB

```{r}
## costs supimab
  total_costs_supi <- sum(c_mat_disc[,c("c_pfs_supi","c_pps_supi")])
  total_costs_supi
  
  ## costs soc
  total_costs_soc <- sum(c_mat_disc[,c("c_pfs_soc","c_pps_soc")])
  total_costs_soc
  
  ## qalys supimab
  total_qalys_supi <-sum(q_mat_disc[,c("q_pfs_supi","q_pps_supi")])
  total_qalys_supi
  
  ## qalys soc
  total_qalys_soc <- sum(q_mat_disc[,c("q_pfs_soc","q_pps_soc")])
  total_qalys_soc
  
  ## ICER
  mean( (total_costs_supi - total_costs_soc) / (total_qalys_supi - total_qalys_soc) )
  
  ## INB at 20k
  inb_20 <-  (total_qalys_supi - total_qalys_soc) * 20000 - (total_costs_supi - total_costs_soc)
  inb_20
  
  ## INB at x
  wtp_steps <- seq(0,50000,1000)
  inb_x <- (total_qalys_supi - total_qalys_soc) * wtp_steps - (total_costs_supi - total_costs_soc)
  kable(cbind(wtp_steps,inb_x), digits = 0)
```
<br><br>

*****

<br><br>
